Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-10-18 article.tex
More specific description of study in introduction
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1411 ccdd05d5-56c1-461a-bf98-85f6876523b4
Shuffled around introduction
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1413 ccdd05d5-56c1-461a-bf98-85f6876523b4

This work will focus on reducing the latency of trigger production.  Data
analysis strategies for advance detection of \CBC{}s will have to strike a
balance between latency and throughput. \CBC\ searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW\ searches.  The most obvious example of the latter is
direct time domain (\TD) convolution, which is latency-free.  However, its cost in floating
point operations per second is linear in the length of the templates, so it is
prohibitively expensive for long templates.  Several efforts are under way to
develop low-latency \CBC\ search pipelines with tractable computing requirements.
One example is \mbta{}~\citep{Marion2004, Buskulic2010}, which was deployed in
S6/VSR3.  MBTA consists of multiple, usually two, template banks for different frequency bands, one which is matched to the early inspiral and the other which is matched to the late inspiral.  An excursion in the output of any filter bank triggers coherent reconstruction of the full matched filtered output.  Final triggers are built from the reconstructed matched filter output.  Another novel approach using networks of parallel, second order IIR filters is being explored
by \citet{shaunIIR, linqingIIR}.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of known low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as the singular value decomposition (\SVD)~\citep{Cannon:2010p10398}
or the Gram-Schmidt process~\citep{rbf} to reduce the number of templates.  We
will use both aspects to demonstrate that a very low latency detection statistic
is possible with current computing resources.  Assuming the other technical
sources of latency can be reduced significantly, this could make it possible to
send prompt alerts to the astronomical community.

The paper is organized as follows.  First, we discuss prospects for early-warning
detection.  Then, we provide an overview of our novel method for detecting compact
binary coalescence signals near real-time. We then describe a prototype
implementation using open source signal processing software.  To validate our approach
we present results of simulations.  We conclude with some remarks on what remains to
prepare for the advanced detector era.

>>>>

This work will focus on reducing the latency of trigger production.  Data
analysis strategies for advance detection of \CBC{}s will have to strike a
balance between latency and throughput. \CBC\ searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW\ searches.  The most obvious example of the latter is
direct time domain (\TD) convolution, which is latency-free.  However, its cost in floating
point operations per second is linear in the length of the templates, so it is
prohibitively expensive for long templates.

Fortunately, the morphology of inspiral signals can be exploited to offset some
of the computational complexity of known low-latency algorithms.  First, the signals
evolve slowly in frequency, so that they can be broken into contiguous
band-limited time intervals and processed at possibly lower sample rates.
Second, inspiral filter banks consist of highly similar templates, admitting
methods such as the singular value decomposition (\SVD)~\citep{Cannon:2010p10398}
or the Gram-Schmidt process~\citep{rbf} to reduce the number of templates.

Several efforts that exploit one or both of these properties are under way to
develop low-latency \CBC\ search pipelines with tractable computing requirements.
One example is \mbta{}~\citep{Marion2004, Buskulic2010}, which was deployed in
S6/VSR3.  MBTA consists of multiple, usually two, template banks for different frequency bands, one which is matched to the early inspiral and the other which is matched to the late inspiral.  An excursion in the output of any filter bank triggers coherent reconstruction of the full matched filtered output.  Final triggers are built from the reconstructed matched filter output.  Another novel approach using networks of parallel, second order IIR filters is being explored
by \citet{shaunIIR, linqingIIR}.

We will use both properties to demonstrate that a very low latency detection statistic
is possible with current computing resources.  Assuming the other technical
sources of latency can be reduced significantly, this could make it possible to
send prompt alerts to the astronomical community.

The paper is organized as follows.  First, we discuss prospects for
early-warning detection.  Then, we provide an overview of our novel
method for detecting compact binary coalescence signals near real-time.
We then describe a prototype implementation using open source signal
processing software.  To validate our approach we present a case study
focusing on a particular subset of the \NS--\NS\ parameter space.  We
conclude with some remarks on what remains to prepare for the advanced
detector era.
